Event History Analyses for psychological time-to-event data: A tutorial in R with examples in Bayesian and frequentist workflows

### 2024-10-20 manuscript/manuscript.tex
update ms.

In experimental psychology, it is standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. Such mean-average comparisons have been the workhorse of experimental psychology over the last century, and have had a substantial impact of theory development and our understanding of the structure of cognition and brain function. However, differences in mean RT conceal when an experimental effect starts, how long it lasts, how it evolves with increasing waiting time, and whether its onset is time-locked to other events (insert REF). Such information is useful not only for interpretation of the effects, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).

@@

Why does this matter for research in psychology? Compared to the aggregation of data across trials, a distributional approach offers the possibility to reveal the timecourse of psychological states. As such, the approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to discriminate between different theoretical accounts of psychological and/or brain-based processes.
For example, the distributions in Example 4 show that the effect starts around 200 ms and is gone by 600 ms. In contrast, in Example 5, the effect starts around 400 ms and is gone by 800 ms. And in the Example 6, the effect reverses around between 500 and 600 ms. What kind of theory or set of theories could account for such effects? Are there new auxiliary assumptions that theories need to adopt? And are there new experiments that need to be run to test the novel predictions that follow from these analyses? As we show later using concrete examples from past experimental data, for many psychological questions this ``temporal states'' information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes as well as adding a relatively under-used dimension to our theory building toolkit.

From a historical perspective, it is worth noting that the development of analytical tools that can estimate or predict when events will occur is not a new endeavour.
Indeed, hundreds of years ago, analytical methods were developed to predict time to death (REFs).
The same logic has been applied to psychological time-to-event data, as previously demonstrated (Panis et al., 2020).
Here, in the paper, we hope to show the value of EHA for knowledge and theory building in cognitive psychology and related areas of research, such as cognitive neuroscience, as well as provide practical tutorials that provide step-by-step code and instructions in the hope that we can enable others to use EHA in a more routine, efficient and effective manner.

@@

In this paper, we focus on a distributional method known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis.
We first provide a brief overview of hazard analysis to orient the reader to the basic concepts that we will use throughout the paper. However, this will remain relatively short, as this has been covered in detail before Singer and Willett (2003), Allison (1982), and Allison (2010), and our primary aim here is to introduce a set of tutorials, which explain \textbf{how} to do such analyses, rather than repeat in any detail \textbf{why} you should do them.

We then provide four different tutorials, each of which is written in the R programming language and publicly available on our Github and the Open Science Framework (OSF) pages, along with all of the other code and material associated with the project. The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply the analyses to their own time-to-event data sets. Each tutorial is provided as an RMarkdown file, so that others can download and adapt the code to fit their own purposes. Additionally, each tutorial is made available as .html file, so that it can be viewed by any web browser, and thus available to those that do not use R.

In Tutorial 1, we illustrate how to process or ``wrangle'' a previously published RT dataset to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 2, we illustrate how one can fit Bayesian multi-level regression hazard models to the data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 3, we illustrate how to fit the same type of regression hazard models in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to EHA. In Tutorial 4, we provide a generalisation of the approach to illustrate one might descriptive statistics when using a more complex design, such as when there are two independent variables.

In summary, even though event history analyses is a widely used statistical tool and there already exist many excellent reviews (REFs) and tutorials (Allison, 2010) on its general use-cases, we are not aware of any tutorials that are aimed at psychological time-to-event data, and which provide worked examples of the key data processing and multi-level regression modelling steps.
Thereofore, our ultimate goal is twofold: first, we want to convince readers of the many benefits of using hazard analysis when dealing with time-to-event data with a focus on psychological time-to-event data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform hazard analysis.

@@

For a comprehensive background context to hazard analysis, we recommend several excellent textbooks (Singer \& Willett, 2003). Likewise, for general introduction to understanding regression equations, we recommend several introductory level textbooks (REFs). Our focus here is not on providing a detailed account of the underlying regression equations, since this topics has been comprehensively covered many times before. Instead, we want to provide an intuition to how EHA works in general as well as in the context of experimental psychology. As such, we only supply regression equations in supplementary materials and then refer to them in the text whenever relevant.

@@

The definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units. Since our focus here is on hazard models that use discrete time units, we describe that approach. After dividing time in discrete, contiguous time bins indexed by t (e.g., t = 1:10 timebins), let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. For example, the first response could occur at 546 ms and it would be in timebin 6 (any RTs from 501 ms to 600). Continuous RT data is treated here as interval-censored data.

Discrete-time EHA focuses on the discrete-time hazard function and the discrete-time survivor function (Figure X). The equations that define both of these functions are reported in supplementary materials (part A). The discrete-time hazard probability gives you the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins. In other words, it reflects the instantaneous likelihood that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. This conditionality in the definition of hazard is what makes the hazard function so diagnostic for studying event occurrence, as an event can physically not occur when it has already occurred before. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event \emph{non}occurrence to obtain the probability that the event occurs after bin t. In other words, the survivor function reflects the likelihood that the event occurs in the future, i.e., in one of the subsequent timebins.

The survivor function can help to qualify or provide context to the interpretation of the hazard function. For example, it can give a sense of how many trials may contribute to that part of the distribution. If each participant completes 100 trials in an experiment, and the survivor function prob of 0.03, then only 3\% of trials remain beyond this point, which in this case would amount to 3 trials. Therefore, the error bars in this part of the distribution would be wider and less precise compared to other parts.

@@

\caption{Hazard and survivor functions}\label{fig:plot2-1}
\end{figure}

@@

In the context of experimental psychology, it is common for participants to be presented with a task that has a right and a wrong answer. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time hazard function can be extended with the discrete-time conditional accuracy function (see equ. X in Supps), which gives you the probability that a response is correct given that it is emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977).

Integrating results between hazard and conditional accuracy functions can be informative for understanding psychological processes. To illustrate, we consider a hypothetical example that is inspired by real data (Panis et al., 2016), but simplified to make the main point clearer (Figure 3). In a standard response priming paradigm, there is a prime stimulus (e.g., an arrow pointing left or right) followed by a target stimulus (another arrow pointing left or right). The prime can then be congruent or incongruent with the target. Figure 3 shows that the early upswing in hazard is equal for both conditions, and that early responses are always correct in .. and always incorrect in the incongruent condtion. Taken together, the results show that for early responses (\textless{} bin 6), responses always follow the prime (and not the target, as instructed). And then for later bins, response hazard is lower in incongruent compared to congruent trials, as \ldots..the prime can be overridden, as both conditions are now always correct. This is interesting because mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. And such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

@@

\caption{Hazard and conditional accuracy}\label{fig:plot2-2}
\end{figure}

@@

In such experimental designs, since the number of trials per condition are spread across bins, it is important to have a relatively large number of trial repetitions per participant and per condition. Accordingly, experiment designs using this approach typically focus on factorial, within-subject designs, in which a large number of observations are made on a relatively small number of participants (so-called small-\emph{N} designs). This approach emphasizes the precision and reproducibility of data patterns at the individual participant level to increase the inferential validity of the design (Baker et al., 2021; Smith \& Little, 2018).

@@

Tutorials 1a and 1b show how to calculate and plot the descriptive statistics when there are one and two independent variables, respectively. Tutorials 2a and 2b illustrate how to use Bayesian multilevel modeling to fit hazard and conditional accuracy models, respectively.
Tutorials 3a and 3b show how to implement, respectively, multilevel models for hazard and conditional accuracy in the frequentist framework.
Additionally, to further simplify the process for other users, the tutorials rely on a set of our own custom functions that make sub-processes easier to autmate, such as data wrangling and plotting functions.

@@

1a. Wrangle raw data and descriptive stats for one independent variable.
1b. Wrangle raw data and descriptive stats for two independent variables.

2a. Bayesian multilevel modeling for h(t)
2b. Bayesian multilevel modeling for ca(t)

3a. Frequentist multilevel modeling for h(t)
3b. Frequentist multilevel modeling for ca(t)

@@

Our data wrangling procedures serve two related purposes. First, we want to summarise and visualise descriptive statistics that relate to our main research questions. Second, we want to produce two different datasets that can each be submitted to different types of inferential modelling approaches. The two types of data structure we label as `person-trial' data (Table 1) and `person-trial-bin' data (Table 2). The `person-trial' data will be familiar to most researchers who record behavioural responses from participants, as it represents the measured RT and accuracy per trial within an experiment. In contrast, the `person-trial-bin' data has a different, more extended structure, which indicates in which bin a response occurred, if at all, in each trial. Therefore, the `person-trial-bin' dataset generates a 0 in each bin until an event occurs and then it generates a 1 to signal an event has occurred. It is worth pointing out that there is no requirement for an event to occur at all (in any bin), as maybe there was no response on that trial or the event occurred after the timewindow of interest. Likewise, when the event occurs in bin 1 there would only be one row of data for that trial.

@@

Next, we can set up the life tables and plots of the discrete-time functions h(t), S(t) and ca(t). To do so using a functional programming approach, one has to nest the data within participants using the group\_nest() function, and supply a user-defined censoring time and bin width to our function ``censor()'', as follows.

@@

The next step is to plot the data using our custom made plotting tool plot\_eha(). When creating the plots, some warning messages will likely be generated, like these:

@@

The warning messages are generated because some bins have no hazard and ca(t) estimates, and no error bars. They can thus safely be ignored.
One can now inspect different aspects, including the life table for a particular condition of a particular subject, and a plot of the different functions for a particular participant. It is important to visually inspect the functions first for each participant, in order to identify possible cheaters (e.g., a flat conditional accuracy function at .5 indicates (s)he was only guessing), outlying individulas, and/or different groups with qualitatively different behavior.

Table 3 shows the life table for condition ``blank'' (no prime stimulus presented) for participant 6 - compare to Figure 1. A life table includes for each time bin, the risk set (number of trials that are event-free at the start of the bin), the number of observed events, and the estimates of h(t), S(t), possibly ca(t), and their estimated standard errors (se). At time point zero, no events can occur and therefore h(t) and ca(t) are undefined.

Figure 4 displays the discrete-time hazard, survivor, and conditional accuracy functions for each prime condition for participant 6. By using discrete-time h(t) functions of event occurrence - in combination with ca(t) functions for two-choice tasks - one can provide an unbiased, time-varying, and probabilistic description of the latency and accuracy of responses based on all trials of any data set.

@@

So far in this paper, we have use a simple experimental design, which involved one condition with two levels. But psychological experiments are often more complex, with crossed factorial designs with more conditions and more than two levels. The purpose of Tutorial 1b, therefore, is to provide a generalisation of the basic approach, which extends to a more complicated design. We felt that this might be useful for researchers in experimental psychology that typically use crossed factorial designs.

To this end, Tutorial 1b illustrates how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of Panis and Schmidt (2016), which includes two independent variables: mask type and prime type. As we use the same functional programming approach as in Tutorial 1a, we simply present the sample-based functions for participant 6 in Figure 8.

Compared to the no-mask condition (column 1 in Figure 8), there is a negative compatibility effect in the hazard and conditional accuracy functions when a (relevant, irrelevant, or lines) mask is present.
T\ldots{}

@@

\subsection{4.3 Tutorial 2a: Fitting Bayesian hazard models}\label{tutorial-2a-fitting-bayesian-hazard-models}

@@

There are several analytic decisions one has to make when fitting a hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable is binary, one has to select a link function (see Supps). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose a specification of the effect of discrete TIME (i.e., the time bin index t). One can choose a general specification (one intercept per bin) or a functional specification, such as a polynomial one (compare model 1 with models 2, 3, and 4 below). We provide relevant example regression formulas in supplementary materials.

@@

To get the posterior distribution of each parameters given the data, we need to specify a prior distribution for each parameter. The middle column of Supplementary Figure 4 shows seven examples of prior distributions on the logit and/or cloglog scales.

While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B in Figure 3 show that specifying such distributions on the logit and cloglog scales leads to rather informative distributions on the original probability (i.e., discrete-time hazard) scale, as most mass is pushed to probabilities of 0 and 1.

@@

{[}{[}Here let's give some intuition on why we would want to setup the model like this{]}{]}

For the first model, we use a general specification of TIME (i.e., one intercept per time bin) for the baseline condition (blank prime), and assume that the effects of prime-target congruency and trial number are proportional and additive, and that the effect of trial number is linear.
Before we fit model 1, we remove unnecessary columns from the data, and specify our priors. In the code of Tutorial 2, this is accomplished as follows.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# remove unnecessary columns before fitting a model}
\NormalTok{M1\_data }\OtherTok{\textless{}{-}}\NormalTok{ ptb\_data }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{select}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{c}\NormalTok{(bl,tr,trial,period, period\_9,d9))  }

\CommentTok{\# Specify priors}
\NormalTok{priors\_M1 }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}0.2,0.71,{-}2.2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{),       }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d6"}\NormalTok{),}
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d7"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d8"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d10"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d11"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d12"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d13"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d14"}\NormalTok{), }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"d15"}\NormalTok{),}
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"skew\_normal({-}1,1,{-}2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{, }\AttributeTok{coef =} \StringTok{"Intercept"}\NormalTok{),}
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"normal(0, 1)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"sd"}\NormalTok{),        }
  \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"lkj(2)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"cor"}\NormalTok{)            }
\NormalTok{)}
\end{Highlighting}
\end{Shaded}

We can now estimate our first Bayesion regression model, as follows.

@@

Estimating model M1 took about 70 minutes on a MacBook Pro (Sonoma 14.6.1 OS, 18GB Memory, M3 Pro Chip).

@@

{[}{[}Here let's give some intuition on why we would want to modify the formula and model features{]}{]}

For the second model, we use a third-order polynomial specification of TIME for the baseline condition (blank prime), and again assume that the effects of prime-target congruency and trial number are proportional and additive, and that the effect of trial number is linear. We first remove unncessary columns and specify our priors.

Estimating model M2 took about 144 minutes.

>>>>

In experimental psychology, it is standard practice to analyse response times (RTs), saccade latencies, and fixation durations by calculating average performance across a series of trials. Such mean-average comparisons have been the workhorse of experimental psychology over the last century, and have had a substantial impact of theory development and our understanding of the structure of cognition and brain function. However, differences in mean RT conceal when an experimental effect starts, how long it lasts, how it evolves with increasing waiting time, and whether its onset is time-locked to other events (Panis, 2020; Panis, Moran, Wolkersdorfer, \& Schmidt, 2020; Panis \& Schmidt, 2016, 2022; Panis, Torfs, Gillebert, Wagemans, \& Humphreys, 2017; Panis \& Wagemans, 2009). Such information is useful not only for interpretation of the effects, but also for cognitive psychophysiology and computational model selection (Panis, Schmidt, Wolkersdorfer, \& Schmidt, 2020).

@@

Why does this matter for research in psychology? Compared to the aggregation of data across trials, a distributional approach offers the possibility to reveal the timecourse of psychological states. As such, the approach permits different kinds of questions to be asked, different inferences to be made, and it holds the potential to discriminate between different theoretical accounts of psychological and/or brain-based processes.
For example, the distributions in Example 4 show that the effect starts around 200 ms and is gone by 600 ms. In contrast, in Example 5, the effect starts around 400 ms and is gone by 800 ms. And in the Example 6, the effect reverses around between 500 and 600 ms. What kind of theory or set of theories could account for such effects? Are there new auxiliary assumptions that theories need to adopt? And are there new experiments that need to be run to test the novel predictions that follow from these analyses? As we show later using concrete examples from past experimental data, for many psychological questions this ``temporal states'' information can be theoretically meaningful by leading to more fine-grained understanding of psychological processes as well as adding a relatively under-used dimension -- the passage of waiting time -- to our theory building toolkit.

From a historical perspective, it is worth noting that the development of analytical tools that can estimate or predict whether and when events will occur is not a new endeavour.
Indeed, hundreds of years ago, analytical methods were developed to predict time to death (REFs).
The same logic has been applied to psychological time-to-event data, as previously demonstrated (Panis et al., 2020).
Here, in the paper, we hope to show the value of EHA for knowledge and theory building in cognitive psychology and related areas of research, such as cognitive neuroscience, as well as provide practical tutorials that provide step-by-step code and instructions in the hope that we can enable others to use EHA in a more routine, efficient and effective manner.

@@

In this paper, we focus on a distributional method for time-to-event data known as discrete-time event history analysis, a.k.a. hazard analysis, duration analysis, failure-time analysis, survival analysis, and transition analysis.
We first provide a brief overview of hazard analysis to orient the reader to the basic concepts that we will use throughout the paper. However, this will remain relatively short, as this has been covered in detail before (Allison, 1982, 2010; Singer \& Willett, 2003), and our primary aim here is to introduce a set of tutorials, which explain \textbf{how} to do such analyses, rather than repeat in any detail \textbf{why} you should do them.

We then provide six different tutorials, each of which is written in the R programming language and publicly available on our Github and the Open Science Framework (OSF) pages, along with all of the other code and material associated with the project. The tutorials provide hands-on, concrete examples of key parts of the analytical process, so that others can apply the analyses to their own time-to-event data sets. Each tutorial is provided as an RMarkdown file, so that others can download and adapt the code to fit their own purposes. Additionally, each tutorial is made available as .html file, so that it can be viewed by any web browser, and thus available to those that do not use R.

In Tutorial 1a, we illustrate how to process or ``wrangle'' a previously published RT + accuracy data set to calculate descriptive statistics when there is one independent variable. The descriptive statistics are plotted, and we comment on their interpretation. In Tutorial 1b we provide a generalisation of this approach to illustrate how one might descriptive statistics when using a more complex design, such as when there are two independent variables. In Tutorial 2a, we illustrate how one can fit Bayesian multi-level regression models to RT-hazard data using the R package brms. We discuss possible link functions, and plot the model-based effects of our predictors of interest. In Tutorial 2b we fit Bayesian multi-level regression models to \emph{timed} accuracy data to perform a micro-level speed-accuracy tradeoff (SAT) analysis, which complements the hazard analysis. In Tutorial 3a, we illustrate how to fit the same type of hazard regression models in a frequentist framework using the R package lme4. We then briefly compare and contrast these inferential frameworks when applied to EHA. In Tutorial 3b, we illustrate how to perform the SAT analysis in a frequentist framework.

In summary, even though event history analyses is a widely used statistical tool and there already exist many excellent reviews (REFs) and tutorials (Allison, 2010) on its general use-cases, we are not aware of any tutorials that are aimed at psychological time-to-event data, and which provide worked examples of the key data processing and multi-level regression modelling steps.
Thereofore, our ultimate goal is twofold: first, we want to convince readers of the many benefits of using hazard analysis when dealing with time-to-event data with a focus on psychological time-to-event data, and second, we want to provide a set of practical tutorials, which provide step-by-step instructions on how you actually perform a discrete-time hazard analysis on time-to-event data, as well as a complementary discrete-time SAT analysis on timed accuracy data.

@@

For a comprehensive background context to hazard analysis, we recommend several excellent textbooks (Singer \& Willett, 2003). Likewise, for general introduction to understanding regression equations, we recommend several introductory level textbooks (REFs). Our focus here is not on providing a detailed account of the underlying regression equations, since this topics has been comprehensively covered many times before. Instead, we want to provide an intuition to how EHA works in general as well as in the context of experimental psychology. As such, we only supply regression equations in the supplementary material and then refer to them in the text whenever relevant.

@@

The definition of hazard and the type of models employed depend on whether one is using continuous or discrete time units. Since our focus here is on hazard models that use discrete time units, we describe that approach. After dividing time in discrete, contiguous time bins indexed by t (e.g., t = 1:10 timebins), let RT be a discrete random variable denoting the rank of the time bin in which a particular person's response occurs in a particular experimental condition. For example, the first response might occur at 546 ms and it would be in timebin 6 (any RTs from 501 ms to 600). Continuous RT data is treated here as interval-censored data.

Discrete-time EHA focuses on the discrete-time hazard function and the discrete-time survivor function (Figure 2). The equations that define both of these functions are reported in the supplementary material (equations 1 and 2 in part A). The discrete-time hazard probability gives you the probability that the event occurs (sometime) in bin t, given that the event does not occur in previous bins. In other words, it reflects the instantaneous likelihood that the event occurs in the current bin, given that it has not yet occurred in the past, i.e., in one of the prior bins. This conditionality in the definition of hazard is what makes the hazard function so diagnostic for studying event occurrence, as an event can physically not occur when it has already occurred before. In contrast, the discrete-time survivor function cumulates the bin-by-bin risks of event \emph{non}occurrence to obtain the probability that the event occurs after bin t. In other words, the survivor function reflects the likelihood that the event occurs in the future, i.e., in one of the subsequent timebins.

The survivor function can help to qualify or provide context to the interpretation of the hazard function. For example, it can give a sense of how many trials contribute to each part of the hazard distribution. If a participant completes 100 trials in an experiment, and the survivor function reaches a probability of 0.03 at the end of timebin (400,500{]}, then only 3\% of trials remain beyond this point, which in this case would amount to 3 trials. Therefore, the error bars in later parts of the hazard function would be wider and less precise compared to earlier parts.

@@

\caption{Hazard and survivor functions}\label{fig:plot2}
\end{figure}

@@

In the context of experimental psychology, it is common for participants to be presented with a task that has a right and a wrong answer. For example, a task may involve choosing between two response options with only one of them being correct. For such two-choice RT data, the discrete-time hazard function can be extended with the discrete-time conditional accuracy function (see equation 5 in part A of the supplementary material), which gives you the probability that a response is correct given that it is emitted in time bin t (Allison, 2010; Kantowitz \& Pachella, 2021; Wickelgren, 1977).

Integrating results between hazard and conditional accuracy functions can be informative for understanding psychological processes. To illustrate, we consider a hypothetical example that is inspired by real data (Panis et al., 2016), but simplified to make the main point clearer (Figure 3). In a standard response priming paradigm, there is a prime stimulus (e.g., an arrow pointing left or right) followed by a target stimulus (another arrow pointing left or right). The prime can then be congruent or incongruent with the target. Figure 3 shows that the early upswing in hazard is equal for both prime conditions, and that early responses are always correct in the congruent condition and always incorrect in the incongruent condition. These results show that for early responses (\textless{} bin 6), responses always follow the prime (and not the target, as instructed). And then for later bins, response hazard is lower in incongruent compared to congruent trials, and all responses emitted in these later bins are correct. This is interesting because mean-average RT would only represent the overall ability of cognition to overcome interference, on average, across trials. And such a conclusion is not supported when the effects are explored over a timeline. Instead, the psychological conclusion is much more nuanced and suggests that multiple states start, stop and possibly interact over a particular temporal window.

@@

\caption{Hazard and conditional accuracy}\label{fig:plot3}
\end{figure}

@@

In such experimental designs, since the number of trials per condition are spread across bins, it is important to have a relatively large number of trial repetitions per participant and per condition. Accordingly, experimental designs using this approach typically focus on factorial, within-subject designs, in which a large number of observations are made on a relatively small number of participants (so-called small-\emph{N} designs). This approach emphasizes the precision and reproducibility of data patterns at the individual participant level to increase the inferential validity of the design (Baker et al., 2021; Smith \& Little, 2018).

@@

Tutorials 1a and 1b show how to calculate and plot the descriptive statistics when there are one and two independent variables, respectively. Tutorials 2a and 2b illustrate how to use Bayesian multilevel modeling to fit hazard and conditional accuracy models, respectively. Tutorials 3a and 3b show how to implement, respectively, multilevel models for hazard and conditional accuracy in the frequentist framework.
Additionally, to further simplify the process for other users, the tutorials rely on a set of our own custom functions that make sub-processes easier to autmate, such as data wrangling and plotting functions (see part B in the supplemental material for a list of the custom functions).

@@

1a. Wrangle raw data and descriptive stats for one independent variable.
1b. Wrangle raw data and descriptive stats for two independent variables.
2a. Bayesian multilevel modeling for h(t)
2b. Bayesian multilevel modeling for ca(t)
3a. Frequentist multilevel modeling for h(t)
3b. Frequentist multilevel modeling for ca(t)

@@

Our data wrangling procedures serve two related purposes. First, we want to summarise and visualise descriptive statistics that relate to our main research questions. Second, we want to produce two different datasets that can each be submitted to different types of inferential modelling approaches. The two types of data structure we label as `person-trial' data (Table 1) and `person-trial-bin' data (Table 2). The `person-trial' data will be familiar to most researchers who record behavioural responses from participants, as it represents the measured RT and accuracy per trial within an experiment. In contrast, the `person-trial-bin' data has a different, more extended structure, which indicates in which bin a response occurred, if at all, in each trial. Therefore, the `person-trial-bin' dataset generates a 0 in each bin until an event occurs and then it generates a 1 to signal an event has occurred in that bin. It is worth pointing out that there is no requirement for an event to occur at all (in any bin), as maybe there was no response on that trial or the event occurred after the time window of interest. Likewise, when the event occurs in bin 1 there would only be one row of data for that trial in the person-trial-bin data set.

@@

Next, we can set up the life tables and plots of the discrete-time functions h(t), S(t), ca(t), and P(t) -- see part A of the supplementary material. To do so using a functional programming approach, one has to nest the data within participants using the group\_nest() function, and supply a user-defined censoring time and bin width to our custom function ``censor()'', as follows.

@@

The next step is to plot the data using our custom function plot\_eha(). When creating the plots, some warning messages will likely be generated, like these:

@@

The warning messages are generated because some bins have no hazard and ca(t) estimates, and no error bars. They can thus safely be ignored.
One can now inspect different aspects, including the life table for a particular condition of a particular subject, and a plot of the different functions for a particular participant. It is important to visually inspect the functions first for each participant, in order to identify possible cheaters (e.g., a flat conditional accuracy function at .5 indicates (s)he was only guessing), outlying individuals, and/or different groups with qualitatively different behavior.

Table 3 shows the life table for condition ``blank'' (no prime stimulus presented) for participant 6 -- compare to Figure 1. A life table includes for each time bin, the risk set (number of trials that are event-free at the start of the bin), the number of observed events, and the estimates of h(t), S(t), possibly ca(t), and their estimated standard errors (se). At time point zero, no events can occur and therefore h(t) and ca(t) are undefined.

Figure 4 displays the discrete-time hazard, survivor, conditional accuracy, and probability mass functions for each prime condition for participant 6. By using discrete-time h(t) functions of event occurrence -- in combination with ca(t) functions for two-choice tasks -- one can provide an unbiased, time-varying, and probabilistic description of the latency and accuracy of responses based on all trials of any data set.

@@

So far in this paper, we have use a simple experimental design, which involved one condition with three levels. But psychological experiments are often more complex, with crossed factorial designs with more conditions and more than three levels. The purpose of Tutorial 1b, therefore, is to provide a generalisation of the basic approach, which extends to a more complicated design. We felt that this might be useful for researchers in experimental psychology that typically use crossed factorial designs.

To this end, Tutorial 1b illustrates how to calculate and plot the descriptive statistics for the full data set of Experiment 1 of Panis and Schmidt (2016), which includes two independent variables: mask type and prime type. As we use the same functional programming approach as in Tutorial 1a, we simply present the sample-based functions for participant 6 in Figure 5.

Compared to the no-mask condition (column 1 in Figure 5), there is a negative compatibility effect in the hazard and conditional accuracy functions when a (relevant, irrelevant, or lines) mask is present.

@@

\subsection{4.3 Tutorial 2a: Fitting Bayesian hazard models to time-to-event data}\label{tutorial-2a-fitting-bayesian-hazard-models-to-time-to-event-data}

@@

There are several analytic decisions one has to make when fitting a hazard model. First, one has to select an analysis time window, i.e., a contiguous set of bins for which there is enough data for each participant. Second, given that the dependent variable is binary, one has to select a link function (see part C in the supplementary material). The cloglog link is preferred over the logit link when events can occur in principle at any time point within a bin, which is the case for RT data (Singer \& Willett, 2003). Third, one has to choose a specification of the effect of discrete TIME (i.e., the time bin index t). One can choose a general specification (one intercept per bin) or a functional specification, such as a polynomial one (compare model 1 with models 2, 3, and 4 below). We provide relevant example regression formulas in part D of the supplementary material.

@@

To get the posterior distribution of each model parameter given the data, we need to specify a prior distribution for each parameter. The middle column of Figure 9 in part E of the supplementary material shows seven examples of prior distributions on the logit and/or cloglog scales.

While a normal distribution with relatively large variance is often used as a weakly informative prior for continuous dependent variables, rows A and B in Figure 9 show that specifying such distributions on the logit and cloglog scales leads to rather informative distributions on the original probability scale, as most mass is pushed to probabilities of 0 and 1. The other rows in Figure 9 show prior distributions on the logit and cloglog scale that we use instead.

@@

When you do not want to make assumptions about the shape of the hazard function, or its shape is not smooth but irregular, you can use a general specification of TIME, i.e., one intercept per time bin. In this first model, we use a general specification of TIME for the selected baseline condition (blank prime), and assume that the effects of prime-target congruency and trial number are proportional and additive, and that the effect of trial number is linear.
Before we fit model 1, we remove unnecessary columns from the data, and specify our priors. In the code of Tutorial 2a, model M1 is specified as follows.

@@

After selecting the binomial family and the cloglog link, the model formula is specified. The fixed effects include 9 dummy variables, the explicit Intercept variable (which represents bin 9 in this example), and the main effects of priming condition and centered trial number. Each of these effects is allowed to vary across individuals.
Estimating model M1 took about 70 minutes on a MacBook Pro (Sonoma 14.6.1 OS, 18GB Memory, M3 Pro Chip).

@@

When the shape of the hazard function is rather smooth, one can fit a more parsemoneous model by using a polynomial specification of TIME. For our second example model, we thus use a third-order polynomial specification of TIME for the baseline condition (blank prime), and again assume that the effects of prime-target congruency and centered trial number are proportional and additive, and that the effect of trial number is linear. The model formula for model M2 looks as follows.
